Topics in Statistical Theory --- Example Sheet 1

### 2020-11-10 modern_statistical_methods/msm_ES2.tex
I'm still alive

% Functions that appear frequently
\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\Int}{Int}
\DeclareMathOperator{\Span}{Span}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\begin{proof}
	By the previous question, all Lasso solutions have the same norm $x$, so the set of Lasso solutions is given by
	\[
	\argmin_{\norm{\beta}_1 = x} \norm{Y - X\beta}_2^2.
	\]
	Let $\hat\beta$ be any fixed Lasso solution, then we find that the entire solution set can also be described as
	\[
	\qty{\hat\gamma \mid \norm{\gamma} = x, \hat\beta - \hat\gamma \in \Nul(X)}. 
	\]
	??
\end{proof}

@@

\begin{proof}
	By the KKT conditions any lasso solution satisfies $\frac1n X\T (Y - X\hat\beta) = \frac1n X\T Y - \frac1n X\T X\hat\beta= \lambda \hat\nu$ for some $\nu$ with $\norm{\hat\nu}_\infty \leq 1$ and $\hat\nu_{\hat S_{\lambda}} = \sign(\hat\beta^L_{\lambda, \hat S_\lambda})$.

	Suppose $\lambda \geq \lambda_\Rm{max}$, and suppose also that $\hat\beta \neq 0$, so $\norm{\hat\nu}_\infty = 1$. Then we find
	\[
	\frac1n \norm{X\T (Y - X\hat\beta)}_\infty = \norm{\lambda\hat\nu} = \lambda  \geq \frac1n \norm{X\T Y}_\infty, 
	\]
	so 
	\[
	\norm{X\T Y - X\T X\hat\beta}_\infty \geq \norm{X\T Y}_\infty. 
	\]
	??
\end{proof}
\end{document}

>>>>

% Functions that appear frequently
\DeclareMathOperator{\sign}{sgn}
\DeclareMathOperator{\Int}{Int}
\DeclareMathOperator{\Span}{Span}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\begin{proof}
	Let $v_1, \dotsc, v_p \subseteq \RR^n$ be the columns of $X$. Because $X$ has centred columns, we know that $v_1, \dotsc, v_p$ lie in the $(n-1)$-dimensional subspace of vectors in $\RR^n$ with mean zero.

	Now, let $\hat\beta$ be any lasso solution. If $\hat\beta = 0$, we are done, so suppose $\hat\beta \neq 0$, and write
	\[
	X\hat\beta = \mqty[ v_1 & \dotsb & v_p] \mqty[ \hat\beta_1 \\ \vdots \\ \hat\beta_p] = \norm{\hat\beta}_1 \mqty[\sign(\hat\beta_1) v_1 & \dotsb & \sign(\hat\beta_p) v_p] \mqty[ \abs{\hat\beta_1} / \norm{\hat\beta}_1\\ \vdots \\ \abs{\hat\beta_p}/\norm{\hat\beta}_1].
	\]
	If we define $S = \qty{\norm{\hat\beta}_1 \sign(\hat\beta_1)v_1, \dotsc, \norm{\hat\beta}_1 \sign(\hat\beta_p)v_p}$, then we see that $X\hat\beta$ can be written as a convex combination of points in $S$. Since the points in $S$ also all have mean zero, by Caratheodory's lemma there exists a vector $\gamma \in \RR^p$ with at most $n$ nonzero coefficients, nonnegative entries, and $\norm{\gamma}_1 \geq 0$ such that
	\[
	X\hat\beta = \norm{\hat\beta}_1 \mqty[\sign(\hat\beta_1) v_1 & \dotsb & \sign(\hat\beta_p) v_p] \gamma = X \qty(\norm{\hat\beta}_1 \mqty[\sign(\hat\beta_1) \gamma_1 \\ \vdots \\ \sign(\hat\beta_p) \gamma_p]) \eqqcolon X\hat\beta'. 
	\]
	Now, we have $X\hat\beta = X\hat\beta'$ and $\norm{\hat\beta}'_1 = \norm{\hat\beta}_1$, which shows that $\hat\beta'$ is a Lasso solution with no more than $n$ nonzero coefficients. 
\end{proof}

@@

\begin{proof}
	We first show that 0 is a lasso solution: we have
	\[
	\frac1n X\T (Y - X0) = \frac1n X\T Y = \lambda \cdot \frac{\frac1n X\T Y}{\lambda} \eqqcolon \lambda\hat\nu, 
	\]
	and since 
	\[
	\norm{\hat\nu}_\infty = \frac{\norm{X\T Y}_\infty / n}{\lambda} \leq \frac{\norm{X\T Y}_\infty / n}{\norm{X\T Y}_\infty /n} \leq 1
	\]
	it follows that 0 is a Lasso solution. By question 9, it follows that any other Lasso solution must also have norm 0, and therefore 0 is the only Lasso solution. 
%	By the KKT conditions any lasso solution satisfies $\frac1n X\T (Y - X\hat\beta) = \frac1n X\T Y - \frac1n X\T X\hat\beta= \lambda \hat\nu$ for some $\nu$ with $\norm{\hat\nu}_\infty \leq 1$ and $\hat\nu_{\hat S_{\lambda}} = \sign(\hat\beta^L_{\lambda, \hat S_\lambda})$. 
%	
%	Suppose $\lambda \geq \lambda_\Rm{max}$, and suppose also that $\hat\beta \neq 0$, so $\norm{\hat\nu}_\infty = 1$. Then we find
%	\[
%	\frac1n \norm{X\T (Y - X\hat\beta)}_\infty = \norm{\lambda\hat\nu} = \lambda  \geq \frac1n \norm{X\T Y}_\infty, 
%	\]
%	so 
%	\[
%	\norm{X\T Y - X\T X\hat\beta}_\infty \geq \norm{X\T Y}_\infty. 
%	\]
%	??
\end{proof}

\begin{question}
	Show that when the columns of $X$ are orthogonal (so necessarily $p \leq n$) and scaled to have $\ell_2$-norm $\sqrt n$, the $k$-th component of the Lasso estimator is given by
	\[
	\hat\beta^L_{\lambda, k} = (\abs{\hat\beta_k^\Rm{OLS}} - \lambda)_+ \sign(\hat\beta_k^\Rm{OLS}). 
	\]
	What is the corresponding estimator if the $\ell_1$ penalty $\norm{\beta}_1$ in the Lasso objective is replaced by the $\ell_0$ penalty $\norm{\beta}_0 \ceq \#\qty{k \mid \beta_k \neq 0}$? 
\end{question}

\begin{proof}
	For simplicity, let $\hat\beta$ denote the proposed Lasso estimator and $\tilde\beta$ the OLS estimator. Since the columns of $X$ are orthogonal, $X$ is injective, so the Lasso estimator is unique (since $X\hat\beta_\lambda^L$ is unique).

	Now, define $\gamma \in \RR^p$ by 
	\[
	\gamma_i = \begin{cases}
		\tilde\beta_i, &\abs{\tilde\beta_i}\leq \lambda, \\ \lambda, &\abs{\tilde\beta_i} > \lambda, \\ -\lambda, &\abs{\tilde\beta_i} < -\lambda
	\end{cases},
	\]
	so that $\hat\beta= \tilde\beta - \gamma$. Then we find
	\[
	\frac1n X\T (Y- X\hat\beta) = \frac1n X\T(Y - X(\tilde\beta - \gamma)) = \frac1n X\T (Y - X\tilde\beta)+ \frac1n X\T X\gamma = \frac1n \gamma = \lambda \frac1{n\lambda}\gamma, 
	\]
	since $Y - X\tilde\beta$ is orthogonal to the columns of $X$ so $X\T(Y - X\tilde\beta) = 0$.

	Now, we have $\norm{\frac1{n\lambda} \gamma}_\infty = \frac1{n\lambda} \norm{\gamma}_\infty \leq \frac1n \leq 1$, and furthermore, if $\hat\beta_i \neq 0$, then it is clear that $\sign(\hat\beta_i) = \sign(\gamma_i)$. This shows that $\hat\beta$ is indeed a Lasso solution.

	Suppose we replace the Lasso objective by the $\ell_0$ penalty. Let $\qty{X_1, \dotsc, X_p}$ be the columns of $X$ and expand them to an orthonormal basis $\qty{X_1, \dotsc, X_n}$ of $\RR^n$. Then we can write
	\[
	\norm{Y - X\beta}_2^2 = \norm{\sum_{i=1}^n P_{X_i} Y - \sum_{i=1}^p \beta_i X_i}_2^2 = \sum_{i=1}^p \qty(\ang{X_i, Y} - \beta_i)^2  + \sum_{j = p+1}^n \ang{X_{j}, Y}^2. 
	\]
	Since the latter sum does not depend on $\beta$, our objective function is equivalent to minimising
	\[
	\sum_{i=1}^p \frac1{2n}\qty(\ang{X_i, Y} - \beta_i)^2 + \lambda \ind_{\beta_i \neq 0}. 
	\]
	Clearly, this is minimised by choosing $\beta_i = \tilde\beta_i$ if $\lambda \leq \frac1{2n} \ang{X_i, Y}^2$ and $\beta_i = 0$ otherwise (note, if $\lambda = \frac1{2n} \ang{X_i, Y}^2$, then choosing $\beta_i =0 $ or $\beta_i = \tilde\beta_i$ gives the same outcome). 
%	If we replace the Lasso objective by the $\ell_0$ penalty, note that with $P_{X_i}$ the projection onto $X_i$, we have
%	\[
%	\norm{Y - X\tilde\beta}_2^2 = \norm{Y}_2^2 - \sum_{i=1}^p \norm{P_{X_i}Y}_2^2,
%	\]
%	so we obtain the $\norm{\cdot}_0$ estimator by replacing $\tilde{\beta_i}$ by 0 if $\norm{P_{X_i} Y} > \lambda$. 
	\end{proof}
\end{document}

### 2020-11-10 topics_statistical_theory/C2.tex
I'm still alive

Putting everything together, we obtain the following:
\begin{theorem}
	Fix $\beta, L > 0$, and let $K$ be a kernel of order $\ell = \ceil\beta$ with $R(k)$ and $\mu_\beta(K)$ finite. Let $\hat f_n$ be the KDE with kernel $K$ and bandwidth $h$ based on $X_1, \dotsc, X_n \iid f \in \FF_\Nn(\beta, L)$. Then we have
	\[
	\MISE\hat f_n \leq \frac{R(K)}{nh} + \frac{L^2}{((\ell - 1)!)^2} \mu_\beta^2(K) h^{2\beta}. 
	\]
	In particular, fixing $\alpha > 0$ and taking $h = \alpha n^{-1/(2\beta + 1)}$, there exists $C > 0$ independent of $n$ such that
	\[
	\sup_{f \in \FF_\Nn(\beta, L)} \MISE \hat f_{n, h, K} \leq C n^{-2\beta/(2\beta + 1)}. 
	\]
\end{theorem}

\subsection{Bandwidth selection}
The choice of bandwidth in the previous theorem is not practical since we have not specified $\alpha$ and $\beta$ is typically unknown. One possible approach is \emph{least squares cross validation}. For this, note that minimising the $\MISE$ is equivalent to minimising
\[
\MISE(\hat f_n) - \int_\RR f^2(x) \dd{x} = \EE\qty[ \int_\RR \hat f_n^2(x) \dd{x}] - 2\EE\qty[ \int_\RR \hat f_n(x) f(x) \dd{x}],
\]
and it can be checked that an unbiased estimator for the above is given by 
\[
\Rm{LSCV}(h) \ceq \int_\RR \hat f_n^2(x) \dd{x} - \frac2n \sum_{i=1}^n \hat f_{n, -i}(X_i),
\]
where $\hat f_{n, -i}$ is the KDE based on all observations except $X_i$. We now choose $h$ such that LSCV is minimised.

>>>>

Putting everything together, we obtain the following:
\begin{theorem} \label{thm:mise_bound}
	Fix $\beta, L > 0$, and let $K$ be a kernel of order $\ell = \ceil\beta$ with $R(k)$ and $\mu_\beta(K)$ finite. Let $\hat f_n$ be the KDE with kernel $K$ and bandwidth $h$ based on $X_1, \dotsc, X_n \iid f \in \FF_\Nn(\beta, L)$. Then we have
	\[
	\MISE\hat f_n \leq \frac{R(K)}{nh} + \frac{L^2}{((\ell - 1)!)^2} \mu_\beta^2(K) h^{2\beta}. 
	\]
	In particular, fixing $\alpha > 0$ and taking $h = \alpha n^{-1/(2\beta + 1)}$, there exists $C > 0$ independent of $n$ such that
	\[
	\sup_{f \in \FF_\Nn(\beta, L)} \MISE \hat f_{n, h, K} \leq C n^{-2\beta/(2\beta + 1)}. 
	\]
\end{theorem}

\subsection{Bandwidth selection}
The choice of bandwidth in the previous theorem is not practical since we have not specified $\alpha$ and $\beta$ is typically unknown.

\subsubsection{Least squares cross validation}
One possible approach is \emph{least squares cross validation}. For this, note that minimising the $\MISE$ is equivalent to minimising
\[
\MISE(\hat f_n) - \int_\RR f^2(x) \dd{x} = \EE\qty[ \int_\RR \hat f_n^2(x) \dd{x}] - 2\EE\qty[ \int_\RR \hat f_n(x) f(x) \dd{x}],
\]
and it can be checked that an unbiased estimator for the above is given by 
\[
\Rm{LSCV}(h) \ceq \int_\RR \hat f_n^2(x) \dd{x} - \frac2n \sum_{i=1}^n \hat f_{n, -i}(X_i),
\]
where $\hat f_{n, -i}$ is the KDE based on all observations except $X_i$. We now choose $h$ such that LSCV is minimised.

\subsubsection{Lepski}
\TODO write this subsection (\TODO understand this first)

\subsection{Choice of kernel}
To choose a kernel, we first fix the scale of the kernel by setting $\mu_2(K) = 1$. Now, by our bound on the MISE (\cref{thm:mise_bound}) it is reasonable to minimise $R(K)$, where for simplicity we assume that $K$ is a nonnegative second-order kernel. The solution is the Epanechnikov kernel
\[
K_E(x) \ceq \frac{3}{4\sqrt 5} \qty(1 - \frac{x^2}{5}) \ind_{\abs{x} \leq \sqrt 5}, 
\]
and the ratio $R(K_E)/R(K)$ is called the \emph{efficiency} of a kernel $K$. We find that for different kernels, the efficiency is greater than 0.9, which shows that kernel shape does not affect efficiency greatly.

\subsection{Multivariate density estimation}
The general $d$-dimensional KDE is 
\[
\hat f_n(x) \ceq \frac{1}{n \sqrt{\det H}} \sum_{i=1}^n K(H^{-1/2}(x - X_i)),
\]
where $H$ is a symmetric positive-definite bandwidth bandwidth matrix (often chosen to be diagonal or a multiple of $I$). If $H = h^2 I$, it can be shown that, under an appropriate definition of a $\beta$ smoothness class, we have an optimal bandwidth of order $n^{-1/(d + 2\beta)}$, and with this choise, a MISE of order $n^{-2\beta/(d + 2\beta)}$. This is called the ``curse of dimensionality'': the higher the dimension becomes, the harder nonparametric estimation gets.
